US National Weather Service Storm Data. Impact on population health and economic consequences
========================================================
# Synopsis  
In this analysis we process the data regarding US severe ewather events that 
can cause public health and economic problems. These events can have a 
significant cost in human fatalities and injuries as well as property damage and
other financial consequences. We start by downloading online the NOOA database 
from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2.
We continue by cleaning our data and aggregating to find the Event Types with
the most important outcomes. We present our results using plots. The analysis
was performed with version 0.98.501 of RStudia on Windows 8 x64



# Data Processing  

### Download initial datafile 
```{r cache=TRUE}
myurl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url=myurl, destfile="StormData.csv.bz2")
```

### Extract csv file and read into a dataframe variable
```{r cache=TRUE}
extracted.file<-bzfile("StormData.csv.bz2")
mydata<-read.csv(extracted.file)
```
### Have a first glance at data
```{r cache=TRUE}
str(mydata)
names(mydata)
```

### For our analysis we are interested only in the following columns:  
EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP
### So we create a new data frame containing only these columns

```{r cache=TRUE}
keep<-c("EVTYPE","FATALITIES", "INJURIES","PROPDMG","PROPDMGEXP","CROPDMG","CROPDMGEXP")
mydata2<-mydata[keep]
names(mydata2)
```

### Now we focus in the EVTYPE column
```{r cache=TRUE}
len<-length(unique(mydata2$EVTYPE))
print(len)
```
### So we have `r len` different event types
### Let's convert column to uppercase
```{r cache=TRUE}
mydata2$EVTYPE<-toupper(mydata2$EVTYPE)
len2<-length(unique(mydata2$EVTYPE))
print(len2)
```
### We see that we managed to reduce the different event types to `r len2`

### Event type has values like **"TORNADO F2"** and **"TORNADO F3"** so we consolidate these values 
```{r cache=TRUE}
mydata2$EVTYPE[grep("TORNADO", mydata2$EVTYPE)] <- "TORNADO"
mydata2$EVTYPE[grep("FLOOD", mydata2$EVTYPE)] <- "FLOOD"
mydata2$EVTYPE[grep("RAIN", mydata2$EVTYPE)] <- "RAIN"
mydata2$EVTYPE[grep("FIRE", mydata2$EVTYPE)] <- "FIRE"
mydata2$EVTYPE[grep("HEAT", mydata2$EVTYPE)] <- "HEAT"
mydata2$EVTYPE[grep("SNOW", mydata2$EVTYPE)] <- "SNOW"
mydata2$EVTYPE[grep("WIND", mydata2$EVTYPE)] <- "WIND"
mydata2$EVTYPE[grep("COLD", mydata2$EVTYPE)] <- "COLD"
mydata2$EVTYPE[grep("THUNDERSTORM", mydata2$EVTYPE)] <- "THUNDERSTORM"
mydata2$EVTYPE[grep("HAIL", mydata2$EVTYPE)] <- "HAIL"
mydata2$EVTYPE[grep("ICE", mydata2$EVTYPE)] <- "ICE"
mydata2$EVTYPE[grep("HURRICANE", mydata2$EVTYPE)] <- "HURRICANE"


len3<-length(unique(mydata2$EVTYPE))
print(len3)
```
### Now we managed to reduce the different event types to `r len3`

### Aggregate by Event type to get sum by injuries and fatalities
```{r cache=TRUE}
library(plyr)
groupByInjuries<-ddply(mydata2,~EVTYPE,summarise,sum=sum(INJURIES))
groupByInjuries<-groupByInjuries[groupByInjuries$sum>0,]
groupByInjuries$EVTYPE<-factor(groupByInjuries$EVTYPE)

groupByFatalities<-ddply(mydata2,~EVTYPE,summarise,sum=sum(FATALITIES))
groupByFatalities<-groupByFatalities[groupByFatalities$sum>0,]
groupByFatalities$EVTYPE<-factor(groupByFatalities$EVTYPE)
```

# Results
### Order the grouped datasets and plot to find the most harmful event types

```{r}
groupByInjuries<-groupByInjuries[order(groupByInjuries$sum,decreasing=T),]
groupByInjuries2<-groupByInjuries[1:10,]
with(groupByInjuries2,plot(EVTYPE,sum,
                          main="Total injuries by event type",
                          col="darkred"))

groupByFatalities<-groupByFatalities[order(groupByFatalities$sum,decreasing=T),]
groupByFatalities<-groupByFatalities[1:10,]
with(groupByFatalities,plot(EVTYPE,sum,
                          main="Total fatalities by event type",
                          col="darkred"))

```

### Therefore tornados have the most impact on fatalities and injuries